---
title: "DATA2902 Report"
author: "Matty Hempstead 490424010"
subtitle: "Analysis of Class Survey"
date: "23/09/2020"
output: 
  html_document:
      code_folding: hide
      df_print: paged
      fig_caption: yes
      number_sections: yes
      self_contained: yes
      toc: yes
      toc_float: yes
---

```{r setup, include=FALSE}
library(readr)


```

# Introduction
A class survey was performed among DATA2X02 students...

```{r, output=F}
data = readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTf8eDSN_2QbMTyVvO5bdYKZSEEF2bufDdYnLnL-TsR8LM-6x-xu1cxmDMohlbLrkMJn9DE7EG7pg5P/pub?gid=1724783278&single=true&output=csv", col_types=cols())

# Rename all columns
colnames(data) = c("timestamp", "covid_test", "gender", "postcode", "dentist", "university_hours", "social_media", "dog_or_cat", "live_with_parents", "exercise_hours", "eye_colour", "asthma", "employment_hours", "favourite_season", "shoe_size", "height", "floss", "glasses_or_contacts", "dominant_hand", "steak_preference", "stress_rating")
```

**Is this a random sample of DATA2X02 students?**
```{r}
cat("Number of samples:", dim(data)[1])
```
There are X people enrolled in DATA2X02, only 174 answered?
It's not a random sample, it is specifically the group of people who decided to answer the form.


**What are the potential biases? Which variables are most likely to be subjected to this bias?**
Many questions rely on memory and/or estimation?
e.g. Estimating hours of work/excersise/paid employment.


**Are there any questions that needed improvement to generate useful data?**
Does useful data include clean data?

Many questions should specify units. Shoe size (AU/UK), height.

Other, more subjective questions could made more objective.
Stress level should mention that 0=not stressed at all and 10=constant high level of stress.
Reason being is some people may interpret values like 10 as referring to a constant high level of stress, where as other may see 10 as being the most stressed they have ever been. 

Hours exercising/studying should be made more specific, similar to the paid employment question, which generally has an objective understanding. 
Should define what is meant by excercise (going for a walk? swim at the beach? jogging/running?).
Also define what is meant by uni work. Does this include watching lectures/attending tuts?


## minor heading

dwadwadwa
wadawd
## minor heading

dwadwadwa
wadawd


# Executive Summary
this is a test

## minor heading

dwadwadwa
wadawd

## minor heading

dwadwadwa
wadawd
## minor heading

dwadwadwa
wadawd


# Executive Summary
this is a test

## minor heading

dwadwadwa
wadawd

## minor heading

dwadwadwa
wadawd
## minor heading

dwadwadwa
wadawd
